Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry

### 2016-11-02 ms_pre_post.Rmd
Make minor tweaks to the text, especially concerning the framing as an empirical paper, not a theory paper; fix spelling errors in multiple places.

Inbreeding avoidance has been widely hypothesised to evolve through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no modelling compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: (1) Does evolution of post-copulatory inbreeding avoidance facilitate evolution of costly polyandry? (2) Is evolution of one costly inbreeding strategy inhibited by another non-costly alternative strategy? And (3) is evolution of one inbreeding strategy precluded by existence of another already adaptive strategy? We show that evolution of polyandry is facilitated by post-copulatory inbreeding avoidance. Evolution of relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in biological systems.

@@

Inbreeding often greatly reduces the fitness of inbred offspring [termed 'inbreeding depression'; @Charlesworth1999; @Keller2002; @Charlesworth2009]. Such strong inbreeding depression is widely hypothesised to drive evolution of inbreeding avoidance, which can be enacted through multiple reproductive decisions in species with obligate biparental reproduction [@Parker1979; @Parker2006; @Pusey1996; @Szulkin2012]. From the perspective of a focal breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), or by biasing fertilisation towards unrelated males after mating has occurred, meaning that related males are less likely to sire offspring (i.e., post-copulatory inbreeding avoidance). Importantly, evolution of post-copulatory inbreeding avoidance requires that females are polyandrous, defined as mating with multiple males during a single reproductive bout. Further, such polyandry could itself evolve because it allows females to mate with additional unrelated males following initial mating with a relative, thereby facilitating pre-copulatory inbreeding avoidance even in the absence of any post-copulatory female choice or otherwise biased fertilisation (e.g., given a 'fair raffle' among sperm provided by feamles' mates). As a reproductive strategy, polyandry can thereby simultaneously allow females to mate with less closely related males and creates the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002; @Akcay2007]. However, despite such well-established hypotheses, no modelling investigates conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are likely to evolve, or predicts how such evolution will feed back to affect underlying evolution of polyandry.

<!--- What is 'quantitative theory'? I don't think that I understand what the implied knowledge gap is here, actually. We set up an empirical system to test some interesting hypotheses; I suppose that the knowledge gap is whether or not those specific hypotheses are supported given the parameters we model. The hypothesis testing sort of relates to the brief verbal theory that we mention in the introduction, but this manuscript does not address any theoretical knowledge gap using modelling. I don't think this is an esoteric issue; we have to make it clear that we know what we're doing and why we're doing it. --->

Meanwhile, evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both, has been found across diverse taxa. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre-copulatory or post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre-copulatory or post-copulatory inbreeding avoidance despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no modelling framework that predicts *a priori* what combinations should be favoured by selection when both pre-copulatory and post-copulatory inbreeding avoidance, and polyandry, can evolve. Such a framework would allow prediction of the form of mating system evolution across different systems that experience inbreeding depression and allow existing studies to be compared to model predictions.

@@

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models thereby isolate the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

@@

Here we use individual-based modelling to set up *in silico* experiments, which test the hypothesis that, given strong inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance facilitates evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution such that only the less costly inbreeding avoidance strategy will be expressed in the long term. Finally, we test the hypothesis that evolution of one inbreeding avoidance strategy (i.e., pre-copulatory or post-copulatory) will be precluded if it is introduced into a population that is already fixed for the other inbreeding avoidance strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across different reproductive systems.

@@

We model evolution of polyandry and pre-copulatory and post-copulatory inbreeding avoidance by tracking reproductive interactions among individual females and males in a small focal population. Individuals are defined as related if they share a common ancestor within the population pedigree (i.e., non-zero kinship). Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of the three reproductive strategy traits: tendency for polyandry ($P_{a}$), pre-copulatory inbreeding avoidance ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding avoidance ($F_{a}$ i.e., 'fertilisation' alleles). Alleles can take any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each trait. In overview, each generation proceeds with females paying costs, expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locas with no physical linkage. Alleles can then mutate and offspring express inbreeding depression. Immigrants arrive in the population and density regulation limits population growth. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating. The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, repectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding strategies as appropriate because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases as a linear function with each trait value.

@@

After costs are paid, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate, and causes some stochastic variation aroudn the expected mean $N_{males}$ of $P_{p}+1$. A female's value of $P_{p}$ depends on the sum of her 20 $P_{a}$ allele values, which might be negative or positive. However, because females cannot mate with a negative number of males, $P_{p}$ is constrained to be zero or positive such that $P_{p} = 0$ when $P_{g} < 0$, but $P_{p} = P_{g}$ when $P_{g} \geq 0$. Polyandry is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie] that is influenced by continuous genetic variation but only expressed at the threshold value of $P_{g} > 0$. Any negative $P_{g}$ value therefore generates phenotypic monandry, while positive $P_{g}$ values can lead to different degrees of realised polyandry (i.e., different numbers of mates, $N_{males}$).

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Here, each female chooses her $N_{males}$ mates based on her pre-copulatory inbreeding strategy phenotype ($M_{p}$), which equals her pre-copulatory inbreeding strategy genotype ($M_{g}$) defined as the sum of the values of her 20 $M_{a}$ alleles. Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire of her offspring $j$ ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability )i.e., hard rather than sof selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We also assume that inbreeding always decreases offspring viability, $\beta > 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial.

@@

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values were substantially higher than in comparable simulations where post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry. However, $P_{a}$ values were expeted to be positive only given a low cost of polyandry ($c_{P} < 0.005$, Figure 1A,B). This implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

Strong post-copulatory inbreeding avoidance consistently evolved in all simulations where such such evolution was allowed (Figure 1B,D,F,H), even when $P_{a}$ values were expected to be slightly negative (i.e., when there is selection against polyandry, Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is commonly expressed (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2). This means that there is commonly opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of $F_{a}$ allele values.

@@

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero and could not evolve, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Evolution of post-copulatory inbreeding avoidance faciliated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

@@

We have used *in silico* experiments to test three key hypotheses concerning evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. The modelling that we develop provides a novel framework for approaching questions concerning inbreeding avoidance, and for predicting evolution of polyandry and inbreeding avoidance through pre-copulatory and post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

We predict evolution of polyandry in populations where inbreeding depression is severe and evolution of inbreeding avoidance through post-copulatory mechanisms is not constrained. Post-copulatory mechanisms of inbreeding avoidance have been observed under these conditions in experimental systems across diverse taxa [e.g. @Pizzari2004; @Firman2008; @Bretman2009; @Gasparini2011; @Tuni2013; @Firman2015]. For example, the number of sperm arriving to eggs is lower for full siblings of female red jungefowl (*Gallus gallus*) than it is for non-relatives, even after controlling for effects of order of oviposition and social familiarity [@Pizzari2004]. Similarly, in house mice, fertilisation was biased toward unrelated males independent of mating order, perhaps due to post-copulatory mechanisms expressed through the secretion of gametic proteins originating in eggs [@Firman2008; @Firman2015]. Interestingly, female house mice also show pre-copulatory inbreeding avoidance  [@Potts1991; @Roberts2003]. Our results suggest that post-copulatory inbreeding avoidance is unlikely to evolve when females can already avoid inbreeding through pre-copulatory mechanisms (Figure 4), meaning that evolution of post-copulatory inbreeding avoidance could reasonably be hypothesised to have preceded evolution of pre-copulatory inbreeding avoidance in mice.

@@

Recent studies have shown how females can successfully bias fertilisation after mating has occurred. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantities of sperm from full-siblings and unrelated males, more eggs were fertilised by unrealted males because the veolcities of full siblings' sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Our results suggest that evolution of such post-copulatory inbreeding avoidance might increase selection for further polyandry, especially if pre-copulatory inbreeding avoidance is costly (Figure 1).

>>>>

Inbreeding avoidance has been widely hypothesised to evolve through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no modelling compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: (1) Does evolution of post-copulatory inbreeding avoidance facilitate evolution of costly polyandry? (2) Is evolution of one costly inbreeding strategy inhibited by another non-costly alternative strategy? And (3) is evolution of one inbreeding strategy precluded by existence of another already adaptive strategy? We show that evolution of polyandry is facilitated by post-copulatory inbreeding avoidance. Evolution of relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for understanding evolution of both pre-copulatory and post-copulatory inbreeding strategies in biological systems.

@@

Inbreeding often greatly reduces the fitness of inbred offspring [termed 'inbreeding depression'; @Charlesworth1999; @Keller2002; @Charlesworth2009]. Such strong inbreeding depression is widely hypothesised to drive evolution of inbreeding avoidance, which can be enacted through multiple reproductive decisions in species with obligate biparental reproduction [@Parker1979; @Parker2006; @Pusey1996; @Szulkin2012]. From the perspective of a focal breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), or by biasing fertilisation towards unrelated males after mating has occurred, meaning that related males are less likely to sire offspring (i.e., post-copulatory inbreeding avoidance). Importantly, evolution of post-copulatory inbreeding avoidance requires that females are polyandrous, defined as mating with multiple males during a single reproductive bout. Further, such polyandry could itself evolve because it allows females to mate with additional unrelated males following initial mating with a relative, thereby facilitating pre-copulatory inbreeding avoidance even in the absence of any post-copulatory female choice or otherwise biased fertilisation (e.g., given a 'fair raffle' among sperm provided by feamles' mates). As a reproductive strategy, polyandry can thereby simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002; @Akcay2007]. However, despite such well-established hypotheses, no modelling investigates conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are likely to evolve, or predicts how such evolution will feed back to affect underlying evolution of polyandry.

<!--- What is 'quantitative theory'? I don't think that I understand what the implied knowledge gap is here, actually. We set up an empirical system to test some interesting hypotheses; I suppose that the knowledge gap is whether or not those specific hypotheses are supported given the parameters we model? The hypothesis testing sort of relates to the brief verbal theory that we mention in the introduction, but this manuscript does not address any theoretical knowledge gap using modelling. I don't think this is an esoteric issue; we have to make it clear that we know what we're doing and why we're doing it. Hence, I think that we need to be clear that we are addressing an empirical knowledge gap derived from verbal theory -- or reframe the key questions. --->

Meanwhile, evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both, has been found across diverse taxa. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre-copulatory or post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre-copulatory or post-copulatory inbreeding avoidance despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no clear understanding of what combinations should be favoured by selection when both pre-copulatory and post-copulatory inbreeding avoidance, and polyandry, can evolve. Empirical hypothesis testing using computer simulation models can address key questions concerning mating system evolution across different systems that experience inbreeding depression, and can allow existing studies to be compared to model outcomes.

@@

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models thereby isolated the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other. For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

@@

Here we use individual-based modelling to set up *in silico* experiments, which test the hypothesis that, given strong inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance facilitates evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution such that only the less costly inbreeding avoidance strategy will be expressed in the long term. Finally, we test the hypothesis that evolution of one inbreeding avoidance strategy (i.e., pre-copulatory or post-copulatory) will be precluded if it is introduced into a population that is already fixed for the other inbreeding avoidance strategy that is adaptive. We thereby provide a framework for understanding evolution of pre-copulatory and post-copulatory inbreeding avoidance across different reproductive systems.

@@

We model evolution of polyandry and pre-copulatory and post-copulatory inbreeding avoidance by tracking reproductive interactions among individual females and males in a small focal population. Individuals are defined as related if they share a common ancestor within the population pedigree (i.e., non-zero kinship). Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of the three reproductive strategy traits: tendency for polyandry ($P_{a}$), pre-copulatory inbreeding avoidance ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding avoidance ($F_{a}$ i.e., 'fertilisation' alleles). Alleles can take any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each trait. In overview, each generation proceeds with females paying costs, expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression. Immigrants arrive in the population and density regulation limits population growth. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating. The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here, $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding strategies as appropriate because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases as a linear function with each trait value.

@@

After costs are paid, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate, and causes some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$. A female's value of $P_{p}$ depends on the sum of her 20 $P_{a}$ allele values, which might be negative or positive. However, because females cannot mate with a negative number of males, $P_{p}$ is constrained to be zero or positive such that $P_{p} = 0$ when $P_{g} < 0$, but $P_{p} = P_{g}$ when $P_{g} \geq 0$. Polyandry is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie] that is influenced by continuous genetic variation but only expressed at the threshold value of $P_{g} > 0$. Any negative $P_{g}$ value therefore generates phenotypic monandry, while positive $P_{g}$ values can lead to different degrees of realised polyandry (i.e., different numbers of mates, $N_{males}$).

@@

Most often, $N_{males}$ will be smaller than the total number of available males [@Duthie]. Here, each female chooses her $N_{males}$ mates based on her pre-copulatory inbreeding strategy phenotype ($M_{p}$), which equals her pre-copulatory inbreeding strategy genotype ($M_{g}$), defined as the sum of the values of her 20 $M_{a}$ alleles. Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire of her offspring $j$ ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability ) i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We also assume that inbreeding always decreases offspring viability, $\beta > 0$ (i.e., no outbreeding depression). Therefore, because $k_{ij}$ is constrained to values between zero and one, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial.

@@

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values were substantially higher than in comparable simulations where post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry. However, $P_{a}$ values were expected to be positive only given a low cost of polyandry ($c_{P} < 0.005$, Figure 1A,B). This implies that, even given strong inbreeding depression in offspring viability, inbreeding avoidance might generally be a weak force driving polyandry evolution.

Strong post-copulatory inbreeding avoidance consistently evolved in all simulations where such evolution was allowed (Figure 1B,D,F,H), even when $P_{a}$ values were expected to be slightly negative (i.e., when there is selection against polyandry, Figure 1F,H). This reflects the threshold nature of phenotypic expression of polygenic polyandry, wherein random sampling of alleles means that polyandry is commonly expressed (i.e., $P_{g}>0$) even when mean $P_{a}$ values are negative (Figure 2). This means that there is commonly opportunity for expression of post-copulatory inbreeding avoidance and associated selection that drives evolution of $F_{a}$ allele values.

@@

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero and could not evolve, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Evolution of post-copulatory inbreeding avoidance facilitated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

@@

We have used *in silico* experiments to test three key hypotheses concerning evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. The modelling that we develop provides a novel framework for approaching complex questions concerning inbreeding avoidance, and for predicting evolution of polyandry and inbreeding avoidance through pre-copulatory and post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

We observed evolution of polyandry in populations where inbreeding depression was severe and evolution of inbreeding avoidance through post-copulatory mechanisms was not constrained. Post-copulatory mechanisms of inbreeding avoidance have been observed under these conditions across diverse taxa [e.g. @Pizzari2004; @Firman2008; @Bretman2009; @Gasparini2011; @Tuni2013; @Firman2015]. For example, the number of sperm arriving to eggs is lower for full siblings of female red jungefowl (*Gallus gallus*) than it is for non-relatives, even after controlling for effects of order of oviposition and social familiarity [@Pizzari2004]. Similarly, in house mice, fertilisation was biased toward unrelated males independent of mating order, perhaps due to post-copulatory mechanisms expressed through the secretion of gametic proteins originating in eggs [@Firman2008; @Firman2015]. Interestingly, female house mice also show pre-copulatory inbreeding avoidance  [@Potts1991; @Roberts2003]. Our results suggest that post-copulatory inbreeding avoidance is unlikely to evolve when females can already avoid inbreeding through pre-copulatory mechanisms (Figure 4), meaning that evolution of post-copulatory inbreeding avoidance might typically precede evolution of pre-copulatory inbreeding avoidance in mice.

@@

Recent studies have shown how females can successfully bias fertilisation after mating has occurred. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantities of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full siblings' sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Our results suggest that evolution of such post-copulatory inbreeding avoidance might increase selection for further polyandry, especially if pre-copulatory inbreeding avoidance is costly (Figure 1).
